Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2022-01-26 publications/frontiers/dora_outlier_detection_frontiers.tex
Standardize dataset ordering and combine results figs. #79

\subsubsection{Sparsity}
Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because the number
of random splits required to isolate the sample is small.

@@

\begin{table}
\caption{Number of samples in the training and test sets for each dataset,
ordered by dataset size.}
\label{tab:datasets}
\centering
\begin{tabular}{l|c|cc}
  \toprule
  Dataset & Training & \multicolumn{2}{c}{Test}\\
   & Unlabeled &  Outliers &  Inliers \\

  \midrule
  Planetary & 992 &  9 & 49 \\
  Earth & 6,757 & 37 & 76  \\
  F-MNIST & 60,000 & 1,000 & 1,000  \\
  Astrophysics & 100,000   &  25,339 &  74,661 \\
  \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Precision at $N=n_{outlier}$ for four datasets; the best result for each data set is in bold.}
\label{tab:faw_results}
\centering
\begin{tabular}{l|cccc}
  \toprule
  Algorithm & Astro. & Earth & Planetary & F-MNIST \\
  \midrule
  PCA               & \textbf{0.42} & 0.41 & 0.44 & \textbf{0.84} \\
  DEMUD             & ---- & 0.41 & 0.44 & \textbf{0.84} \\
  \midrule
  RX                & 0.40 & 0.43 & ---- & 0.82 \\
  LRX               & ----& ---- & 0.33 & 0.56 \\
  \midrule
  Iso. Forest  & 0.34 & 0.46 & \textbf{0.56} & 0.74 \\
  \midrule
  PAE               & 0.35 & 0.41 & 0.44 & 0.83 \\
  Neg. Sampling & 0.32 & \textbf{0.49} & 0.33 & 0.43 \\
  \midrule
  Random            & 0.25 & 0.32 & 0.14 & 0.50 \\
  \bottomrule
\end{tabular}
\end{table}

@@

% TODO: Tried to make these a 2x2 panel of subfigures, but was getting
% errors using the subfigure package (Error: 'Not in outer par mode'). It seems
% this option may not be available for Frontiers articles 
% (see bottom of document). May combine them as a single figure manually.

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/planetary_combined_plot.pdf}
\caption{Number of known outliers ranked in top $N$ selections for 
planetary rover dataset.}
\label{fig:planetary_results}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/faw_combined_plot.pdf}
\caption{Number of known outliers ranked in top $N$ selections for 
Earth science dataset.}
\label{fig:faw_results}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/des_combined_plot.pdf}
\caption{Number of known outliers ranked in top $N$ selections for 
the astrophysics (DES) dataset.}
\label{fig:des_results}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/fmnist_combined_plot.pdf}
    \caption{Number of known outliers ranked in top $N$ selections for 
    F-MNIST dataset.}
    \label{fig:fmnist_results}
\end{figure}

@@

%\todo{Hannah: Consider merging all three data set discussions since they are
%in one table? with their own figures? The point here isn't really to
%compare algs or data sets but instead to show that we CAN compare
%them, using DORA.}
% We initialized each algorithm with $D_s$ and requested a ranking of
% targets in $D_{s+}$.
%

For the planetary dataset, we found that
the Isolation Forest achieved the highest precision (best outlier
detection) when allowed to select only \num{9} images.
Figure~\ref{fig:planetary_results} shows the complete (cumulative)
outlier detection performance for each algorithm when ranking
all \num{58} target images in $D_{1694-1703}$. We could not
employ RX since the data dimensionality ($64 \times 64 =
4096$) exceeded the data set size.

For the Earth dataset, negative sampling had the best performance
in both metrics. DEMUD, PCA, and PAE tied for the lowest precision at 
$N=n_{outlier}$ while DEMUD and PCA tied for the lowest
MDR (Figure \ref{fig:faw_results}). We did not evaluate LRX for this time 
series dataset because LRX can only be applied to gridded image or raster 
data types.

For the astrophysics dataset (Figure~\ref{fig:des_results}), DEMUD was omitted due to computational time 
and LRX was omitted as it applies only to image data.  Of the remaining methods, 
PCA achieved the highest precision, followed by RX.  Negative sampling 
performs well initially before its performance drops off.  The PAE finds the most outliers 
overall.

For the MNIST and F-MNIST dataset, PCA and DEMUD tied for the highest
precision at $N=n_{outlier}$
while DEMUD, PCA, and PAE tied for the highest MDR 
(Figure \ref{fig:fmnist_results}). 
Negative sampling had the 
lowest performance in both metrics.

>>>>

\subsubsection{Sparsity}
Isolation forest (iForest) is a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because the number
of random splits required to isolate the sample is small.

@@

\begin{table}
\caption{Number of samples in the training and test sets for each dataset.}
\label{tab:datasets}
\centering
\begin{tabular}{l|c|cc}
  \toprule
  Dataset & Training & \multicolumn{2}{c}{Test}\\
   & Unlabeled &  Outliers &  Inliers \\
  \midrule
  Astrophysics & 100,000   &  25,339 &  74,661 \\
  Planetary & 992 &  9 & 49 \\
  Earth & 6,757 & 37 & 76  \\
  F-MNIST & 60,000 & 1,000 & 1,000  \\
  \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Precision at $N=n_{outlier}$ for four datasets; the best result for each data set is in bold.}
\label{tab:faw_results}
\centering
\begin{tabular}{l|cccc}
  \toprule
  Algorithm & Astro. & Planetary & Earth & F-MNIST \\
  \midrule
  PCA               & \textbf{0.42} & 0.44 & 0.41 & \textbf{0.84} \\
  DEMUD             & --- & 0.44 & 0.41 & \textbf{0.84} \\
  \midrule
  RX                & 0.40 & --- & 0.43 & 0.82 \\
  LRX               & --- & 0.33 & --- & 0.56 \\
  \midrule
  iForest  & 0.34 & \textbf{0.56} & 0.46 & 0.74 \\
  \midrule
  PAE               & 0.35 & 0.44 & 0.41 & 0.83 \\
  Neg. Sampling & 0.32 & 0.33 & \textbf{0.49} & 0.43 \\
  \midrule
  Random            & 0.25 & 0.14 & 0.32 & 0.50 \\
  \bottomrule
\end{tabular}
\end{table}

@@

% Tried to make these a 2x2 panel of subfigures, but was getting
% errors using the subfigure package (Error: 'Not in outer par mode'). It seems
% this option may not be available for Frontiers articles 
% (see bottom of document). May combine them as a single figure manually.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/combined_results.pdf}
\caption{Number of known outliers ranked in top $N$ selections for 
the (a) astrophysics, (b) planetary, (c) Earth, and (d) FMNIST datasets.}
\label{fig:combined_results}
\end{figure}

%\begin{figure}
%\centering
%\includegraphics[width=0.6\linewidth]{figures/des_combined_plot.pdf}
%\caption{Number of known outliers ranked in top $N$ selections for 
%the astrophysics (DES) dataset.}
%\label{fig:des_results}
%\end{figure}
%
%\begin{figure}
%\centering
%\includegraphics[width=0.6\linewidth]{figures/planetary_combined_plot.pdf}
%\caption{Number of known outliers ranked in top $N$ selections for 
%planetary rover dataset.}
%\label{fig:planetary_results}
%\end{figure}
%
%\begin{figure}
%\centering
%\includegraphics[width=0.6\linewidth]{figures/faw_combined_plot.pdf}
%\caption{Number of known outliers ranked in top $N$ selections for 
%Earth science dataset.}
%\label{fig:faw_results}
%\end{figure}
%
%\begin{figure}
%    \centering
%    \includegraphics[width=0.6\linewidth]{figures/fmnist_combined_plot.pdf}
%    \caption{Number of known outliers ranked in top $N$ selections for 
%    F-MNIST dataset.}
%    \label{fig:fmnist_results}
%\end{figure}

@@

For the astrophysics dataset (Figure~\ref{fig:combined_results}a), 
DEMUD was omitted due to computational time 
and LRX was omitted as it applies only to image data.  
Of the remaining methods, 
PCA achieved the highest precision, followed by RX.  Negative sampling 
performs well initially before its performance drops off.  
The PAE finds the most outliers overall.

For the planetary dataset, we found that
the Isolation Forest achieved the highest precision (best outlier
detection) when allowed to select only \num{9} images.
Figure~\ref{fig:combined_results}b shows the complete (cumulative)
outlier detection performance for each algorithm when ranking
all \num{58} target images in $D_{1694-1703}$. We could not
employ RX since the data dimensionality ($64 \times 64 =
4096$) exceeded the data set size.

For the Earth dataset, negative sampling had the best performance
in both metrics. DEMUD, PCA, and PAE tied for the lowest precision at 
$N=n_{outlier}$ while DEMUD and PCA tied for the lowest
MDR (Figure \ref{fig:combined_results}c). We did not evaluate LRX for this time 
series dataset because LRX can only be applied to gridded image or raster 
data types.

For the MNIST and F-MNIST dataset, PCA and DEMUD tied for the highest
precision at $N=n_{outlier}$
while DEMUD, PCA, and PAE tied for the highest MDR 
(Figure \ref{fig:combined_results}d). 
Negative sampling had the 
lowest performance in both metrics.
